The watershed transform in ITK - discussion and new developments

### 2006-05-29 article/Article.tex
Article updates and gel segmentation code
darcs-hash:20060529003507-fafb9-322bfc10e63dd03b7bfa6b3de09b8a9d4048ce10.gz

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-rgb}}
\subfigure[Minima level = 10]{\includegraphics[scale=3.2]{level10-rgb}}
\subfigure[Minima level = 20]{\includegraphics[scale=3.2]{level20-rgb}}
\subfigure[Minima level = 30 and Minima level = 40]{\includegraphics[scale=3.2]{level30-rgb}}
\subfigure[Minima level = 50]{\includegraphics[scale=3.2]{level50-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markersMf-rgb}}
\caption{Different results of morphological watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image. All the watersheds were performed with $MarkWatershed = true$ and $FullyConnected = false$.\label{fig:artificial-itk}}
\end{center}
\end{figure}

@@

\subsection{???}

\section{Discussion}
There is a level of erroneous folklore surrounding the watershed
transform and it is hoped that this report helps to clear the air.

\section{Related work}
This work have lead us to develop some other filters to:
\begin{itemize}
  \item Find regional extrema \cite{regionalExtrema}
  \item Surimpose a colored label image onto an image \cite{labelOverlay}
  \item Impose the minima \cite{minimaImposition}
  \item Label the connected component \cite{connectedComponent}
  \item Invert the intensity of an image \cite{invert}
  \item Enhance the morphological reconstruction \cite{recons}
  \item Compute some informations about the shape of the labeled objects (not published yet)
\end{itemize}

>>>>

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-rgb}}
\subfigure[Minima level = 10]{\includegraphics[scale=3.2]{level10-rgb}}
\subfigure[Minima level = 20]{\includegraphics[scale=3.2]{level20-rgb}}
\subfigure[Minima level = 30 and Minima level = 40]{\includegraphics[scale=3.2]{level30-rgb}}
\subfigure[Minima level = 50]{\includegraphics[scale=3.2]{level50-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markersMf-rgb}}
\caption{Different results of morphological watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image. All the watersheds were performed with $MarkWatershedLine = true$ and $FullyConnected = false$.\label{fig:artificial-itk}}
\end{center}
\end{figure}

@@

\subsection{2D electrophoresis gel segmentation}
Protein separation in two dimensions on electrophoresis gels is one of
the many situations where image segmentation techniques are commonly
used to make measurements of spots which are often touching. This
section describes a multi-stage watershed approach and illustrates the
use of the watershed transform on gradient images and the use of
watershed lines, which are a new capability of the filters introduced
in this paper, as markers. This technnique almost certainly isn't the
best way of segmenting gels, and probably won't work on many gels
without tuning. It is simply offered as an example of how watershed
transforms may be used. There are many variants of the basic
algorithm, some of which are discussed in the text.

The algorithm source code is in {\tt segGel.cxx} and the images used
in this example are from \url{http://www-lecb.ncifcrf.gov/flicker/}.

The original image is shown in Figure \ref{fig:origGel}. The spots we
are interested in are dark, and frequently touching. In gels often
have all sorts of artifacts, ranging from streaks along separation
dimensions to an over abundance of a protein that leads to a large
dark spot that may saturate the image brightness (i.e be darker than
the dark floor of the imaging system). We will not address these sorts
of issues here.

The basic steps are
\begin{itemize}
\item {\bf Find spot markers.} Here we'll use a combination of smoothing 
and minima detection. Depending on the noise levels of the image there
are likely to be minima in bright regions of the image. It may be
possible to discard a number of minima based on brightness, depending
on how certain we are about the maximum brightness of a valid spot.

The results of an unmodified minima detection after median smoothing
is shown in Figure \ref{fig:minimaGel}.
\item {\bf Find background marker.} We'll use a watershed on the raw image 
to generate a background marker. The background marker will be the
watershed line. Setting attribute \verb$MarkWatershedLine$ to
\verb$true$ will result in the watershed line being marked with zero 
values. In this situation we are using the watershed to tesselate the
image into regions. The region boundaries will be in the brightest
parts of the image between markers. Figure \ref{fig:backgroundGel} 
illustrates the results of this step.
\item {\bf Combine markers.} The spot and background markers are combined 
by adding them together, after using the binary threshold filter to
generate the background marker from the first watershed -- see Figure \ref{fig:combinedGel}.
\item {\bf Compute gradient.} A gradient of the raw image is then computed. 
We use the {\tt GradientRecursiveGaussianImageFilter}, but many
options are available. The gradient image is shown in Figure
\ref{fig:gradientGel}.
\item {\bf Compute watershed transform} using the gradient image as the 
control surface. Figure \ref{fig:resultGel}.
\item {\bf Optional additional watershed steps.} The results of this sort 
of segmentation, based on an image gradient, often appear to
underestimate the spot size. This is because the watershed transform
will produce edges along image ridges, which correspond to the
locations of maximum image gradient. Unfortunately this tends not to
correspond to the location a human observer would select.

This easy easy to understand if a gaussian spot is considered -- the
position of maximal gradient is halfway down the edge of the spot,
while a human would tend to mark the edge as being near the point
where the spot brightness becomes indistinguishable from the
background brightness. This location is often closer to the position
of the maximum of the second derivative.

Therefore we can ``improve'' the segmentation by using the previous
results to develop a new set of markers and apply the watershed
transform using the second derivative as a control surface. It is
sensible to use a different set of spot markers to the original
because there are two maxima of second derivative, and the original
markers will tend to be inside both, while the results of the first
stage will be contain the closer maxima. Figures
\ref{fig:combinedGel2}, \ref{fig:gradientGel2} and \ref{fig:resultGel2}
illustrate these steps.

\item {\bf Postprocessing.} This implementation doesn't do any post 
processing, but in most cases it would be a useful step. Error
situations, such as spots being a crazy size, shape or brightness
could be detected brought to the attention of the user or attempts
could be made to automatically correct the result.
\end{itemize}
\begin{figure}[htbp]
\begin{center}
\subfigure[Input]{\includegraphics[scale=0.4]{leukemia-ALL-Human-b01633-Flicker-NCI}\label{fig:origGel}}
\subfigure[Regional minima]{\includegraphics[scale=0.4]{gel_minima}\label{fig:minimaGel}}
\subfigure[Background marker]{\includegraphics[scale=0.4]{gel_tesselation}\label{fig:backgroundGel}}

\subfigure[All markers]{\includegraphics[scale=0.4]{gel_allmarkers}\label{fig:combinedGel}}
\subfigure[Gradient]{\includegraphics[scale=0.4]{gel_gradient}\label{fig:gradientGel}}
\subfigure[Result (background marker removed)]{\includegraphics[scale=0.4]{gel_result}\label{fig:resultGel}}
\subfigure[All markers - stage 2]{\includegraphics[scale=0.4]{gel_allmarkers2}\label{fig:combinedGel2}}
\subfigure[Second derivative (gamma corrected)]{\includegraphics[scale=0.4]{gel_gradient2_gamma2}\label{fig:gradientGel2}}
\subfigure[Result stage 2 (background marker removed)]{\includegraphics[scale=0.4]{gel_result2}\label{fig:resultGel2}}

\caption{Steps used in segmenting a gel using multiple watershed stages.}
\end{center}
\end{figure}

A cursory examination of the results will show that segmentation is
far from perfect. Some touching spots only have one marker so only one
spot is found and there are markers placed in what should probably be
condsidered as background. These problems may be addressed in a number
of ways, such as performing more careful selection of regional minima
and sharpening the original image using deconvolution, as well as
various post processing options. We are not going to explore these
issues here because the aim is to illustrate a few ways in which the
watershed transform can be used.

\section{Discussion}
There is a level of erroneous folklore surrounding the watershed
transform and it is hoped that this report helps to clear the air, as
well as providing a useful tutorial/introduction to a useful tool.

\section{Related work}
This work have lead us to develop some other filters to:
\begin{itemize}
  \item Find regional extrema \cite{regionalExtrema}
  \item Superimpose a colored label image onto an image \cite{labelOverlay}
  \item Impose the minima \cite{minimaImposition}
  \item Label the connected component \cite{connectedComponent}
  \item Invert the intensity of an image \cite{invert}
  \item Enhance the morphological reconstruction \cite{recons}
  \item Compute some informations about the shape of the labeled objects (not published yet)
\end{itemize}
